---
title: "Creating the ``r params$package_name`` R package"
author: "Jacob Bien"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
knit: litr::render
output: litr::litr_gitbook
params:
  package_name: "flowkernel" # <-- change this to your package name
  package_parent_dir: "." # <-- relative to this file's location
documentclass: book
---

# Preliminaries

This document uses [litr](https://jacobbien.github.io/litr-project/) to define the ``r params$package_name`` R package.  When the `index.Rmd` file is rendered, the R package is created along with the bookdown you are reading.  To do so in RStudio, you can simply open  `index.Rmd` and press "Knit" to render the bookdown (and open `_book/index.html` to see the result).  More generally, in a console you can run the following:

```{r, eval = FALSE}
litr::render("index.Rmd")
```

## DESCRIPTION file

We start by specifying some basic information for the description file:

```{r package-setup, message=FALSE, results='hide'}
usethis::create_package(
  path = ".",
  fields = list(
    Package = params$package_name,
    Version = "0.0.0.9000",
    Title = "Smoothly Varying Mixture of Gaussians Modeling",
    Description = "This package uses kernel-smoothed EM to estimate a smoothly varying mixture of Gaussians model.",
    `Authors@R` = person(
      given = "Jacob",
      family = "Bien",
      email = "jbien@usc.edu",
      role = c("aut", "cre")
      )
  )
)
usethis::use_mit_license(copyright_holder = "F. Last")
```

## Package-level documentation

Let's include some package-level documentation. Besides being user-friendly, it's also needed because we'll be using "import from" later.

```{package_doc}
#' Smoothly Varying Mixture of Gaussians Modeling
#'
#' This package uses kernel-smoothed EM to estimate a smoothly varying mixture of Gaussians model.
#' 
#' @docType package
```
